Value Ordering, Discrepancies, and Restarts for Subgraph Algorithms

### 2018-10-23 paper/paper.tex
Tweaks

The subgraph isomorphism problem is to decide whether a copy of a small ``pattern'' graph occurs
inside a larger ``target'' graph. Although \NP-complete, by combining design techniques from
artificial intelligence with careful algorithm engineering, modern subgraph isomorphism solvers can
often produce exact solutions quickly even on graphs with thousands of vertices. The current single
strongest subgraph isomorphism solver uses ``highest degree first'' as a value-ordering heuristic to
direct a constraint programming style search
\cite{DBLP:conf/cp/McCreeshP15,DBLP:conf/lion/KotthoffMS16,DBLP:journals/jair/McCreeshPST18}. This
heuristic is much better than branching randomly, but is still far from perfect. We therefore
investigate different techniques for offsetting mistakes in value-ordering decisions. We first try
discrepancy searches, which turn out to be much too expensive to be a practical solution. We then
look at using random value-ordering, restarts, and nogood recording, which together give better
aggregate performance at the expense of much worse behaviour on many instances. Next, we introduce a
new way of adapting the existing value-ordering heuristic to introduce just a small amount of
randomness.  This lets us make a state-of-the-art subgraph algorithm perform much better
on a large number of satisfiable instances, whilst performing worse only rarely on satisfiable
instances, and never on unsatisfiable instances. Finally, we show that this strategy is also
effective in an optimisation setting, producing benefits in two maximum common induced subgraph
algorithms. This combination of techniques gives us, for the first time, a practical alternative to
backtracking search for constraint programming style algorithms where we have a strong
value-ordering heuristic and where we care both about satisfiable and unsatisfiable instances.

@@

\Cref{figure:scatter-heuristics} gives a detailed look at these results. This scatter plot shows
that despite the aggregate performance being similar, on a case by case basis, the two heuristics
can make a large difference to the performance of satisfiable instances.

@@

\Cref{figure:scatter-dds} shows the effects of adding Walsh's \cite{DBLP:conf/ijcai/Walsh97}
depth-bounded discrepancy search (DDS) to \cref{algorithm:sip} (results with other discrepancy
search variants are similar). Each point represents the solving
time for one instance---to avoid noise for easier instances, we measure only time spent during
search, and exclude time spent in preprocessing and initialisation.  Points below the $x-y$ diagonal
are speedups, whilst points on the top and right axes represent instances which timed out after one
thousand seconds with one algorithm, but not the other. For satisfiable instances, the different
point styles show the different families, whilst all unsatisfiable instances are shown as dark dots.

The points well below the diagonal line and along the right-hand axis show that DDS can sometimes be
extremely beneficial on satisfiable instances.  However, on both unsatisfiable and most satisfiable
instances, the overheads can be extremely large, and DDS is much worse in aggregate and is not a
viable approach (even when only considering satisfiable instances).  These large overheads are to be
expected: discrepancy searches are aimed primarily at getting better feasible solutions in
optimisation problems which are too large for a proof of optimality to be a realistic prospect, and
they are not well-suited for unsatisfiable decision problems. Despite this, the extremely large gains on
some satisfiable instances confirm our suspicions that we should find an alternative to
heavy-commitment backtracking search.

@@

Another alternative to plain backtracking search is provided by \emph{restarts}. The general idea is
to perform a certain amount of search, and then if no solution has been found (and unsatisfiability
has not been proven), to abandon search and restart from the beginning. Obviously, such an approach
can only be beneficial if something changes after restarting---in a constraint programming setting,
this is usually the variable-ordering heuristic
\cite{DBLP:journals/jsat/LecoutreSTV07,DBLP:conf/cp/GayHLS15,DBLP:conf/aaai/LeeSZ16,DBLP:conf/cp/GlorianBLLM17}.
In this paper, we instead rely upon randomness in a special new \emph{value}-ordering heuristic, and continue
to use smallest domain first with static tiebreaking for variable-ordering.\footnote{It may be
possible to further improve \cref{algorithm:sip} by also introducing randomness or some form of
learning into its variable-ordering heuristic. However, simultaneously introducing a second change
would considerably complicate the empirical analysis.} Using restarts on value-ordering heuristics
appears to be uncommon---although \citet{DBLP:conf/flairs/RazgonOP07} look at learning
value-ordering heuristics from restarts, \citet{DBLP:conf/cp/ChuSS09} use a similar
scheme in the context of parallel search, and an early approach by \citet{DBLP:conf/aaai/GomesSK98} does so in an optimisation context.

@@

We use the Luby scheme \cite{DBLP:journals/ipl/LubySZ93} to determine when to restart. We count the
number of backtracks (that is, when we reach the end of the main $\KwSty{foreach}$ loop on
\siplineref{countrestarts}) to decide when to restart, rather than counting recursive calls.
Following established wisdom, we multiply each item in the Luby sequence by a magic constant---we
used the SMAC automatic parameter tuner \cite{DBLP:conf/lion/HutterHL11} to select the value 660.
(We also tried a geometric restart scheme, which gave less favourable results.)

@@

@@

    \caption{The cumulative number of satisfiable instances solved over time, with and without
    restarts and nogood recording.}
    \label{figure:restarts}
\end{figure}

@@

Having looked at subgraph isomorphism in detail, we now briefly discuss the more general maximum
common induced subgraph problem.  Two recent algorithms for this problem also make use of
backtracking search with degree as a value-ordering heuristic. The k${\downarrow}$ algorithm
\cite{DBLP:conf/aaai/HoffmannMR17} attempts to solve the problem by first trying to solve the
induced subgraph isomorphism problem, and then if that fails, retries allowing a single unmatched
vertex (and thus using weaker invariants), and so on. Due to its similarity to \cref{algorithm:sip},
we can introduce the same bias and restart strategy, requiring only small changes
to handle wildcards in nogoods.

@@

We have shown that it is both possible and beneficial to introduce a small amount of randomness into
backtracking search algorithms for hard subgraph problems, without compromising existing search
order heuristics, and without paying a penalty for unsatisfiable instances. In principle it is
well-known that introducing some form of randomness \cite{DBLP:conf/aaai/GomesSK98} or reduced
commitment \cite{DBLP:conf/ijcai/HarveyG95} into backtracking search can be beneficial, but until
now it has not been clear how to do so efficiently in practice, particularly when we already have
strong search ordering heuristics and when are expecting to encounter large numbers of unsatisfiable
instances. Our results show that the key to success is a combination of aggressive restarts, nogood
recording, and biasing an existing value-ordering heuristic with just a small amount of randomness.

>>>>

The subgraph isomorphism problem is to decide whether a copy of a small ``pattern'' graph occurs
inside a larger ``target'' graph. Although the problem is \NP-complete, by combining design techniques from
artificial intelligence with careful algorithm engineering, modern subgraph isomorphism solvers can
often produce exact solutions quickly even on graphs with thousands of vertices. The current single
strongest subgraph isomorphism solver uses ``highest degree first'' as a value-ordering heuristic to
direct a constraint programming style search
\cite{DBLP:conf/cp/McCreeshP15,DBLP:conf/lion/KotthoffMS16,DBLP:journals/jair/McCreeshPST18}. This
heuristic is much better than branching randomly, but is still far from perfect. We therefore
investigate different techniques for offsetting mistakes in value-ordering decisions. We first try
discrepancy searches, which turn out to be much too expensive to be a practical solution. We then
look at using random value-ordering, restarts, and nogood recording, which together give better
aggregate performance at the expense of much worse behaviour on many instances. Next, we introduce a
new way of adapting the existing value-ordering heuristic to introduce just a small amount of
randomness.  This lets us make a state-of-the-art subgraph algorithm perform much better
on a large number of satisfiable instances, whilst performing worse only rarely on satisfiable
instances, and never on unsatisfiable instances. Finally, we show that this strategy is also
effective in an optimisation setting, producing benefits in two maximum common induced subgraph
algorithms. This combination of techniques gives us, for the first time, a practical alternative to
backtracking search for constraint programming style algorithms where we have a strong
value-ordering heuristic and where we care both about satisfiable and unsatisfiable instances.

@@

\Cref{figure:scatter-heuristics} gives a detailed look at these results.  Each point represents the
solving time for one instance---to avoid noise for easier instances, we measure only time spent
during search, and exclude time spent in preprocessing and initialisation.  Points below the $x-y$
diagonal are speedups, whilst points on the top and right axes represent instances which timed out
after one thousand seconds with one algorithm, but not the other. For satisfiable instances, the
different point styles show the different families, whilst all unsatisfiable instances are shown as
dark dots.  This plot shows that despite the aggregate performance being similar, on a case by case
basis, the two heuristics can make a large difference to the performance of individual satisfiable
instances. This suggests that although degree is a good heuristic, we should perhaps not commit
heavily to a vertex of highest degree, but also consider vertices of the same or similar degree.

@@

\Cref{figure:scatter-dds} shows the effects of adding Walsh's \cite{DBLP:conf/ijcai/Walsh97}
depth-bounded discrepancy search (DDS) to \cref{algorithm:sip} (results with other discrepancy
search variants are similar).  The points well below the diagonal line and along the right-hand axis
show that DDS can sometimes be extremely beneficial on satisfiable instances.  However, on both
unsatisfiable and most satisfiable instances, the overheads can be extremely large, and DDS is much
worse in aggregate and is not a viable approach (even when only considering satisfiable instances).
These large overheads are to be expected: discrepancy searches are aimed primarily at getting better
feasible solutions in optimisation problems which are too large for a proof of optimality to be a
realistic prospect, and they are not well-suited for unsatisfiable decision problems. Despite this,
the extremely large gains on some satisfiable instances confirm our suspicions that we should find
an alternative to heavy-commitment backtracking search.

@@

Another alternative to plain backtracking search is provided by \emph{restarts}. The general idea is
to perform a certain amount of search, and then if no solution has been found (and unsatisfiability
has not been proven), to abandon search and restart from the beginning. Such an approach
can only be beneficial if something changes after restarting---in a constraint programming setting,
this is usually the variable-ordering heuristic
\cite{DBLP:journals/jsat/LecoutreSTV07,DBLP:conf/cp/GayHLS15,DBLP:conf/aaai/LeeSZ16,DBLP:conf/cp/GlorianBLLM17}.
In this paper, we instead rely upon randomness in a special new \emph{value}-ordering heuristic, and continue
to use smallest domain first with static tiebreaking for variable-ordering.\footnote{It may be
possible to further improve \cref{algorithm:sip} by also introducing randomness or some form of
learning into its variable-ordering heuristic. However, simultaneously introducing a second change
would considerably complicate the empirical analysis.} Using restarts on value-ordering heuristics
is uncommon---although \citet{DBLP:conf/flairs/RazgonOP07} look at learning
value-ordering heuristics from restarts, \citet{DBLP:conf/cp/ChuSS09} use a similar
scheme in the context of parallel search, and an early approach by \citet{DBLP:conf/aaai/GomesSK98} does so in an optimisation context.

@@

We use the Luby scheme \cite{DBLP:journals/ipl/LubySZ93} to determine when to restart. We count the
number of backtracks (that is, when we reach the end of the main $\KwSty{foreach}$ loop on
\siplineref{countrestarts}) to decide when to restart, rather than counting recursive calls.
Following convention, we multiply each item in the Luby sequence by a constant---we
used the SMAC automatic parameter tuner \cite{DBLP:conf/lion/HutterHL11} to select the value 660.

@@

    \caption{The number of satisfiable instances solved over time, with and without
    restarts and nogood recording.}
    \label{figure:restarts}
\end{figure}

@@

@@

Having looked at subgraph isomorphism in detail, we now briefly discuss the maximum common induced
subgraph problem, to see whether our new approach to search has more general applicability.  Two
recent algorithms for this problem also make use of backtracking search with degree as a
value-ordering heuristic. The k${\downarrow}$ algorithm \cite{DBLP:conf/aaai/HoffmannMR17} attempts
to solve the problem by first trying to solve the induced subgraph isomorphism problem, and then if
that fails, retries allowing a single unmatched vertex (and thus using weaker invariants), and so
on. Due to its similarity to \cref{algorithm:sip}, we can introduce the same bias and restart
strategy, requiring only small changes to handle wildcards in nogoods.

@@

We have shown that it is both possible and beneficial to introduce a small amount of randomness into
backtracking search algorithms for hard subgraph problems, without compromising existing search
order heuristics, and without paying a penalty for unsatisfiable instances. In principle it is
well-known that introducing some form of randomness \cite{DBLP:conf/aaai/GomesSK98} or reduced
commitment \cite{DBLP:conf/ijcai/HarveyG95} into backtracking search can be beneficial, but until
now it has not been clear how to do so efficiently in practice, particularly when we already have
strong search ordering heuristics and when are expecting to encounter large numbers of unsatisfiable
instances. Our results show that the key to success is a combination of aggressive restarts, nogood
recording, and biasing an existing value-ordering heuristic with just a pinch of randomness.
